---
title: "Model temporal gene expression patterns"
author: "Haotian Zhuang"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Model temporal gene expression patterns}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
date: '`r format(Sys.Date(), "%B %d, %Y")`'
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

# Introduction
We will demonstrate how to use PreTSA to fit the gene expression along pseudotime values and identify temporally variable genes (TVGs).

# Install and load the package
```{r eval=FALSE}
if (!require("devtools"))
install.packages("devtools")
devtools::install_github("haotian-zhuang/PreTSA")
```

```{r setup}
library(PreTSA)
```

# Load and preprocess the dataset
The 3K peripheral blood mononuclear cells (PBMC3K) dataset was downloaded from the [10x website](https://www.10xgenomics.com/resources/datasets/3-k-pbm-cs-from-a-healthy-donor-1-standard-1-1-0). The processed PBMC3K dataset was downloaded using the R package `SeuratData`. `TSCAN` was used to construct a pseudotime trajectory from naive CD4 T cells to memory CD4 T cells with the top 10 principal components (PCs) and the cell clusters obtained from the dataset. Genes that are expressed in fewer than 5 cells within the trajectory were filtered out. We analyzed a final set of 10,509 genes on 1,180 cells.
```{r eval=FALSE}
suppressPackageStartupMessages(library(Seurat))
suppressPackageStartupMessages(library(SeuratData))
suppressPackageStartupMessages(library(TSCAN))
d <- LoadData("pbmc3k", type = "pbmc3k.final")
d <- UpdateSeuratObject(object = d)
dr <- Embeddings(d, reduction = "pca")[, 1:10]
clu <- Idents(d)
order <- guided_tscan(dr = dr, clu = clu, cluorder = c("Naive CD4 T", "Memory CD4 T"))
pseudotime <- setNames(1:length(order), order)
expr <- d@assays$RNA$data
expr <- expr[, names(pseudotime)]
expr <- expr[rowSums(expr > 0) >= 5, ]
```

The processed dataset can be loaded directly from the package to save time.
```{r}
load(system.file("extdata", "sc_pbmc3k", "preprocessed.Rdata", package = "PreTSA"))
str(expr)
str(pseudotime)
```

# Fit the gene expression along pseudotime values
```{r}
fitRes <- temporalFit(expr = expr, pseudotime = pseudotime)
str(fitRes)
```

The argument "knot" indicates the number of knots (0 by default) or automatic selection ("auto"). The argument "maxknotallowed" is the user-defined maximum number of knots (10 by default).
```{r eval=FALSE}
temporalFit(expr = expr, pseudotime = pseudotime, knot = "auto", maxknotallowed = 10)
```

# Identify temporally variable genes (TVGs)
It returns a data frame with the p-value, FDR, test statistic and number of knots selected for each gene. Genes are ordered first by p-value (increasing) then by test statistic (decreasing).
```{r}
testRes <- temporalTest(expr = expr, pseudotime = pseudotime)
head(testRes)
```

To account for the pseudotime uncertainty, we apply the strategy used in `PseudotimeDE`, by subsampling 80\% of cells and using the same procedure as in pseudotime inference on the original dataset.
```{r eval=FALSE}
set.seed(2025)
n.permute <- 100
pseudotime_list <- lapply(1:n.permute, function(i) {
  cell.subsample <- sample(colnames(d), 0.8 * ncol(d), replace = FALSE)
  d.subsample <- subset(d, cells = cell.subsample)
  d.subsample <- RunPCA(d.subsample, npcs = 10, verbose = FALSE)
  dr <- Embeddings(d.subsample, reduction = "pca")[, 1:10]
  clu <- Idents(d.subsample)
  order <- guided_tscan(dr = dr, clu = clu, cluorder = c("Naive CD4 T", "Memory CD4 T"))
  pseudotime <- setNames(1:length(order), order)
  pseudotime
})
pseudotime_permute <- lapply(pseudotime_list, function(i) {
  names(i) <- sample(names(i), replace = FALSE)
  i
})
```

The argument "pseudotime_permute" is a list of permuted pseudotime values from subsampled cells. Each element in the list has the same format of the argument "pseudotime".
```{r}
testRes <- temporalTest(expr = expr[1:100, ], pseudotime = pseudotime, pseudotime_permute = pseudotime_permute)
head(testRes)
```

# Use binning strategy for large datasets
To mitigate memory demands in generating the full gene expression matrix for extremely large datasets, PreTSA can divide the genes into multiple bins (10 by default) and then combine the fitted gene expression matrices obtained from each bin.
```{r}
fitRes <- temporalFit_bin(expr = expr, pseudotime = pseudotime, numbin = 10)
str(fitRes)
```

```{r}
testRes <- temporalTest_bin(expr = expr, pseudotime = pseudotime, numbin = 10)
head(testRes)
```

# Visualize the fitted expression
As an example, *IL32* is an known gene associated to T cell activation process. The fitted PreTSA curve shows an increasing trend.
```{r}
suppressPackageStartupMessages(library(ggplot2))
genename <- "IL32"
df.gene = data.frame(ptime = pseudotime, ori = expr[genename, ], pretsa = fitRes[genename, ])

ggplot(data = df.gene) + geom_point(aes(x = ptime, y = ori), color = "orange") +
  geom_line(aes(x = ptime, y = pretsa), linewidth = 2, color = "royalblue") +
  labs(x = "Pseudotime", y = "Expression", title = genename) + theme_classic() +
  theme(plot.title = element_text(hjust = 0.5, face = "italic"))
```

<details>
  <summary>**Session Info**</summary>
```{r}
sessionInfo()
```
</details>
